---
title: "Spatial Bayesian models"
author: "[Andrei Wong Espejo](https://github.com/Andrei-WongE)"
date: "April 16, 2026"
format:
  html:
    theme: lux
    html-math-method: mathjax
    embed-resources: true
  pdf:
    documentclass: article 
# Add table of contents
    toc: true
    toc_float: true
    toc-depth: 2
    toc-location: left
    toc-title: "On this page"
    number-sections: true
# By default smooth scroll is not enabled
    smooth-scroll: true
    link-external-icon: false
    link-external-newwindow: true
    highlight-style: github
# Fold code (initially hidden)
    code-fold: true
    code_download: yes
    code-link: true
    code-block-bg: true
    code-block-border-left: "#ADBBCB"
    table-captions: true 
    tbl-cap-location: margin
    reference-location: margin
    citation-location: margin
# Text for footer
    include-after-body: footer.html
    
citation:
  type: post
  url: "https://github.com/Andrei-WongE/spatial_bayes_sandbox"
  
# Allow for highlighting text. \hl{highlight} text inline.Or \hl{\textbf{highlight}} with bold face.
# Allows fort commenting using hypothesis
comments:
  hypothesis: true 

execute:
  warning: false
  message: false
  #freeze: auto  # re-render only when source changes
---

# Set-up

```{r}
#| label: timestamp-setup
#| echo: false

knitted_when <- format(Sys.Date())
knitted_where <- knitr::current_input()
knitted_with <- packageVersion("knitr")
knitted_doc_url <- downlit::autolink_url("knitr::knit()")
```

Note updated on `r knitted_when` from `r knitted_where` with knitr version `r knitted_with`.

::: callout-note
## General set-up of the program.

These notes make use of the following R packages and general set-up.

```{r}
#| label: set-up

  if (!require("pacman")) {
  install.packages("pacman")
  }

  pacman::p_load(tidyverse 
                 ,tidylog
                 ,Cairo
                 ,here
                 ,dplyr
                 ,htmltools
                 ,SpatialEpi # for pennLC data
                 ,crsuggest  # Suggest CRS information for spatial data
                 ,ggplot2
                 ,ggnewscale # For multiple fill and colour scales in ggplot2
                 ,sf         # DONT'T FORGET uses the s2geometry library
                 ,leaflet    # JavaScript libraries for interactive maps
                 ,leafem     # Extensions for leaflet 
                 ,knitr
                 ,ragg       # Graphic devices based on the AGG library
                 ,downlit    # For syntax highlighting and automatic linking
                 ,git2r      # Provides access to 'Git' repositories
                 ,xfun       # For alternative Session Info
                 )
  
  #pacman::p_update()         # Update out-of-date packages

  options(scipen = 100, digits = 2)             # Prefer non-scientific notation
  
  knitr::opts_chunk$set( dev       = "ragg_png"
                        ,fig.path  = "Figs/"
                        ,dpi       = 600
                        ,fig.width = 10
                        ,fig.hight = 12
  # Automatically formatting code using styler
                        ,tidy      = "styler"
)

```
:::

# Spatial Bayesian models

## CARBayes Demo

This example is adapted from the following article by [Dr Paula Moraga](https://journal.r-project.org/archive/2018/RJ-2018-036/RJ-2018-036.pdf)

```{r}
#| label: data
#| include: false

source(here("Install.R")) # Has fillmap.R as dependency

# dependencies ‘Rgraphviz’, ‘graph’ are not available
library(INLA)
library(R2OpenBUGS)
library(CARBayes)
library(nimble)

# library(SpatialEpi)  # for pennLC [rdocumentation](https://www.rdocumentation.org/packages/SpatialEpi/versions/1.2.8/topics/pennLC)


```

```{r}
data(pennLC)
names(pennLC)
head(pennLC$data)
head(pennLC$smoking)
```
We calculate the expected number of cases in each county using indirect standardisation. The expected counts in each county represent the total number of disease cases one would expect if the population in the county behaved the way the Pennsylvania population behaves.

```{r}
# Add observed cases
pennLC$data <- pennLC$data %>%
  tibble() %>%
  group_by(county) %>%
  mutate(obs = sum(cases)) %>%
  ungroup()

# Add expected cases
pennLC$data <- pennLC$data %>%
  group_by(race, gender, age) %>%
  mutate(
    pop.strata   = sum(population),
    cases.strata = sum(cases)
  ) %>%
  group_by(county) %>%
  mutate(
    expected = population * cases.strata / pop.strata,
    expected = sum(expected)
  ) %>%
  ungroup()

# Summarise observed and expected; add SIR
pennLC$data <- pennLC$data %>%
  dplyr::select(county, obs, expected) %>%
  distinct() %>%
  mutate(SIR = obs / expected) %>%
  left_join(pennLC$smoking, by = "county")
```

```{r}
# Create sf object from SpatialPolygons
map <- st_as_sf(pennLC$spatial.polygon) %>%
  st_transform("+proj=longlat +datum=WGS84")

map <- cbind(map, pennLC$data)

pal <- colorNumeric(palette = "YlOrRd", domain = map$SIR)
```

```{r}
# Observed SIR map
leaflet(map) %>%
  addTiles() %>%
  addPolygons(
    color       = "grey",
    weight      = 1,
    fillColor   = ~pal(SIR),
    fillOpacity = 0.7
  ) %>%
  addLegend(
    pal      = pal,
    values   = ~SIR,
    opacity  = 0.7,
    title    = "SIR",
    position = "bottomright"
  )
```

```{r}
# Labels
labels <- sprintf(
  "<strong>%s</strong><br/>Observed: %s <br/>Expected: %s 
   <br/>Smokers proportion: %s <br/>SIR: %s",
  map$county,
  map$obs,
  round(map$expected, 2),
  map$smoking,
  round(map$SIR, 2)
) %>%
  lapply(htmltools::HTML)

leaflet(map) %>%
  addTiles() %>%
  addPolygons(
    color           = "grey",
    weight          = 1,
    fillColor       = ~pal(SIR),
    fillOpacity     = 0.7,
    highlightOptions = highlightOptions(weight = 4),
    label           = labels,
    labelOptions    = labelOptions(
      style    = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto"
    )
  ) %>%
  addLegend(
    pal      = pal,
    values   = ~SIR,
    opacity  = 0.7,
    title    = "SIR",
    position = "bottomright"
  )
```
Now we can examine the map and see which areas have

    SIR = 1 which indicates observed cases are the same as expected,
    SIR > 1 which indicates observed cases are higher than expected,
    SIR < 1 which indicates observed cases are lower than expected.

This map gives a sense of the disease risk across Pennsylvania. However, observed (raw) SIRs may be misleading and insufficiently reliable in counties with small populations. In contrast, model-based approaches enable the incorporation of covariates and borrow information from neighbouring counties to improve local estimates, resulting in the smoothing of extreme values based on small sample sizes.

We will use the Besag-York-Mollié (BYM) model first introduced in Besag et al. (1991).

The model simply states that nearby things are similar.

$$ Y_i \sim N(\mu_i, \sigma^2) $$

$$ \mu\_i = z_i \beta + u_i + v_i $$

Spatial random effect ($u_i$) accounting for spatial dependence and modelled with an intrinsic conditional autoregressive model (ICAR) that smooths the data according to a specified neighbourhood structure (for example, using `poly2nb()` to create a neighbours list based on contiguous areas).

The term ($v_i$) is an unstructured exchangeable component modelling uncorrelated noise, assumed i.i.d. normal with zero mean and variance $v_i \sim N(0, \sigma_v^2)$:

ICAR specification for the spatial effect:

$$ u_i \mid u_{-i} \sim N\left(\bar{u}_{\delta_i}, \frac{\sigma_u^2}{n_{\delta_i}}\right) $$

where ($delta_i$) is the set of neighbours of area (i), and ($n_{\delta\_i}$) is the number of neighbours of area (i). The local mean of neighbours is

$$ \bar{u}_{\delta_i} = \frac{1}{n_{\delta_i}} \sum_{j \in \delta_i} u_j $$

Bayesian hierarchical models can be fitted using a number of approaches such as integrated nested Laplace approximation (INLA) (Rue, Martino, and Chopin 2009) and Markov chain Monte Carlo (MCMC) methods (Gelman et al. 2013).

Critiques to the BYM: 

- As indicated by Wu & Banerjee (2024), Simpson et al. (2017) and Riebler et al. (2016) the error components are not identifiable and scaling issues complicate interpretation of the variance parameters in the BYM model.

::: {.callout-note appearance="simple"}
## Pay Attention EXPLAIN BETTER
:::

Set-up:

The BYM model combines two random effects per area: a spatially structured CAR component and an unstructured i.i.d. component. The formula `obs ~ 1 + offset(log(expected))` models observed counts against expected counts, which is standard disease mapping setup yielding relative risk estimates.

MCMC settings `M.burnin = 10000` discards the first 10,000 iterations as the chain converges. `M = 5000` keeps 5,000 posterior samples, making total iterations: 15,000.

```{r}
# Adjacency matrix, converting to binary numeric matrix
W <- st_touches(map) %>%
  as.matrix() * 1

M.burnin <- 10000
M        <- 5000

set.seed(111)

MCMC <- S.CARbym(
  formula  = obs ~ 1 + offset(log(expected)),
  data     = map,
  family   = "poisson",
  W        = W,
  burnin   = M.burnin,
  n.sample = M.burnin + M,
  verbose  = FALSE
)

print(MCMC$summary.results)
```
To obtain the modelled SIR, we need to divide the fitted values by the expected values.

We use `RR` to denote relative risk, which is the smoothed SIR (specifically, the median of the smoothed SIR), to distinguish it from the observed SIR. The posterior probability of the relative risk being greater than 1 (i.e. the proportion of MCMC iterations where the smoothed SIR is greater than 1) is also included, denoted by `PP` in the map area labels.

```{r}
# Posterior SIR and summary
y.fit <- MCMC$samples$fitted
SIR   <- t(t(y.fit) / map$expected)

map$SIR.50  <- apply(SIR, 2, median)
map$SIR.025 <- apply(SIR, 2, quantile, 0.025)
map$SIR.975 <- apply(SIR, 2, quantile, 0.975)
map$PP      <- apply(SIR, 2, function(x) length(which(x > 1))) / M

labels <- sprintf(
  "<strong>%s</strong> <br/>Observed: %s <br/>Expected: %s <br/>
   Smoker proportion: %s <br/>SIR: %s <br/>RR: %s (%s, %s) <br/>PP: %s",
  map$county,
  map$obs,
  round(map$expected, 2),
  map$smoking,
  round(map$SIR, 2),
  round(map$SIR.50, 2),
  round(map$SIR.025, 2),
  round(map$SIR.975, 2),
  round(map$PP, 3)
) %>%
  lapply(htmltools::HTML)

leaflet(map) %>%
  addTiles() %>%
  addPolygons(
    color           = "grey",
    weight          = 1,
    fillColor       = ~pal(SIR.50),
    fillOpacity     = 0.7,
    highlightOptions = highlightOptions(weight = 4),
    label           = labels,
    labelOptions    = labelOptions(
      style    = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto"
    )
  ) %>%
  addLegend(
    pal      = pal,
    values   = ~SIR.50,
    opacity  = 0.7,
    title    = "Relative Risk",
    position = "bottomright"
  )
```

```{r}
#| label: reproducibility 

		    ## Datetime
		    Sys.time()
		
		    ## Repository
		    git2r::repository()
		
		    ## session info
		    xfun::session_info()

```
